Scan-Centric, Frequency-Based Method for Characterizing Peaks from Direct Injection Fourier transform Mass Spectrometry Experiments

### 2022-04-15 doc/peakcharacterization_manuscript.Rmd
getting all to artifacts

Within this combined high-throughput with direct injection context, we previously developed methods to remove high-peak-density artefacts [@mitchellNewMethodsIdentify2018], annotate peaks with assignments using SMIRFE [@mitchellSmallMoleculeIsotope2019], and generate lipid classifications for spectral peak assignments [@mitchellDerivingLipidClassification2020].
We applied all of these methods together in detecting and assigning differential lipids in non-small-cell lung carcinoma (NSCLC) [@mitchellUntargetedLipidomicsNonSmall2021].

@@

::: {custom-style="MDPI_2.2_heading2"}
### Mitigation of High Peak Density Artefacts
:::

@@

::: {custom-style="MDPI_3.1_text"}
We have previously described the presence of high peak density (HPD) artefacts in FT-MS spectra [@mitchellNewMethodsIdentify2018].
Ideally, the peak characterization procedure should reduce and mitigate their presence and effect in the resultant reported peaks.
Their presence should be minimized by removing noise peaks, and removing peak regions that have multiple reported peaks from the same scan.
However, we expect they may also present as characterized peaks that have larger than expected frequency level standard deviations (FSD) when calculated across scans.
These peaks can be detected by simply examining the distribution of FSDs and marking outlier peaks.
To verify the mitigation of HPD regions, we converted centroided m/z's from XCalibur to frequency values using the previously calculated model for that sample, and measured peak density to detect HPD regions, and compare them with the scan-centric peaks and FSD outliers.
XCalibur peak density was measured using a sliding window ten points wide and a stride of one point.

@@

::: {custom-style="MDPI_3.1_text"}
Having a scan-centric workflow for generating the peak centroids means that we have opportunities to evaluate the scans as a whole, as well as the peaks across scans.
For example, we can mark peaks with unusually high frequency standard deviations (FSD) and mitigate the presence of high-peak-density artefacts (see **Mitigation of High Peak Density Artefacts**).
We also sometimes find that from initial peak region that has been marked as a single region, still results in two peaks being characterized in a single scan.
This results in that peak being removed from that scan entirely.

@@

::: {custom-style="MDPI_3.1_text"}
High-peak-density (HPD) artefacts [@mitchellNewMethodsIdentify2018] present as groups of singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations.
Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and peaks with FSDs greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are marked.
The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work to allow comparisons between it and the use of the FSD.
For HPD detection, the peaks in excel output from Xcalibur were used after converting the m/z peak centers into frequency space.
Sliding windows that are 1000 frequency points wide with a stride of 100 points were used for the density calculations.
:::

>>>>

Within this combined high-throughput with direct injection context, we previously developed methods to remove high-peak-density artifacts [@mitchellNewMethodsIdentify2018], annotate peaks with assignments using SMIRFE [@mitchellSmallMoleculeIsotope2019], and generate lipid classifications for spectral peak assignments [@mitchellDerivingLipidClassification2020].
We applied all of these methods together in detecting and assigning differential lipids in non-small-cell lung carcinoma (NSCLC) [@mitchellUntargetedLipidomicsNonSmall2021].

@@

::: {custom-style="MDPI_2.2_heading2"}
### Mitigation of High Peak Density Artifacts
:::

@@

::: {custom-style="MDPI_3.1_text"}
We have previously described the presence of high peak density (HPD) artifacts in FT-MS spectra [@mitchellNewMethodsIdentify2018].
Ideally, the peak characterization procedure should reduce and mitigate their presence and effect in the resultant reported peaks.
Their presence should be minimized by removing noise peaks, and removing peak regions that have multiple reported peaks from the same scan.
However, we expect they may also present as characterized peaks that have larger than expected frequency level standard deviations (FSD) when calculated across scans.
These peaks can be detected by simply examining the distribution of FSDs and marking outlier peaks.
To verify the mitigation of HPD regions, we converted centroided m/z's from XCalibur to frequency values using the previously calculated model for that sample, and measured peak density to detect HPD regions, and compare them with the scan-centric peaks and FSD outliers.
XCalibur peak density was measured using a sliding window ten points wide and a stride of one point.

@@

::: {custom-style="MDPI_3.1_text"}
Having a scan-centric workflow for generating the peak centroids means that we have opportunities to evaluate the scans as a whole, as well as the peaks across scans.
For example, we can mark peaks with unusually high frequency standard deviations (FSD) and mitigate the presence of high-peak-density artifacts (see **Mitigation of High Peak Density Artifacts**).
We also sometimes find that from initial peak region that has been marked as a single region, still results in two peaks being characterized in a single scan.
This results in that peak being removed from that scan entirely.

@@

::: {custom-style="MDPI_3.1_text"}
High-peak-density (HPD) artifacts [@mitchellNewMethodsIdentify2018] present as groups of singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations.
Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and peaks with FSDs greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are marked.
The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work to allow comparisons between it and the use of the FSD.
For HPD detection, the peaks in excel output from Xcalibur were used after converting the m/z peak centers into frequency space.
Sliding windows that are 1000 frequency points wide with a stride of 100 points were used for the density calculations.
:::
